The shape of kelp: environmental influences on kelp morphology

### 2018-11-08 chapter_2.Rmd
some minor house cleaning and changes

@@

In Figure 3 the propagation of a typical offshore wave spectrum is given as produced from a single time-step in SWAN. Figure 3 is presented to clarify the averaged wave exposure maps presented in Figure 2. Tracing the wave height contours into False Bay its clear why this Bay’s western periphery is predominantly sheltered. It should be mentioned that some of the annual winter frontal depression systems pass the Cape Peninsula from the west to east, resulting in wave propagating towards the continent from much more southerly directions. This results in positive and negative wave exposure anomalies all around the peninsula.

@@

The temperature parameters around the Western Cape coast vary among sites and seasons. During February and August, the west side of Cape point experiences decreased temperatures (Fig. 2).  This region also experiences lower mean and maximum temperatures when compared to False Bay, a region known as east of Cape Point. The range of temperatures within False Bay are larger in winter (August) compared to summer (February) (Fig. 3). The sites north of Kommetjie on the west side display larger temperatures ranges than sites south of Kommetjie on the west side.

@@

Not only do the temperature parameters display seasonal fluctuations, but we see similar observations for wave parameters. The west side of Cape Point exhibit increased median direction, SD and mean significant wave height compared to within False Bay (Fig. 4). Sites within False Bay however display increased standard deviations of wave period. Median wind direction is northerly in winter (Fig. 5), and turns southerly in summer.

@@

#### _Laminaria pallida_
Lamina length for laminaria pallida showed no geographical pattern moving from west to east. Kommetjie, Olifantsbos and Batsata Rock showed great variability in lamina length, and both Buffels Bay and Betty’s Bay were visually different to Miller’s Point and Roman Rock, when comparing summary data (Fig.6). Lamina thickness showed great variation across sites, with Baboon Rock, Miller’s Point, A-Frame and Roman rock displaying large lamina thickness, significantly different to the rest of the sites. Lamina weight was observed to vary for Baboon Rock and Betty’s Bay. Neighbouring sites A-Frame and Roman Rock also showed visual difference when comparing boxplot summary statistics. An increase in the number of digits was observed as one moved from Cape Point north along the western side of False Bay. This ceased at Batstata Rock, which exhibited significantly less digits compared to the previous site, Bordjies reef North. Stipe diameter showed some geographical grouping, with west of Cape Point sites exhibiting larger stipe diameters compared to False Bay sites.

@@

For frond length, frond mass, stipe circumferences, stipe length and total length of _Ecklonia maxima_ morphometrics, we see a gradual increase in value as one moves south from St. Helena Bay to Kommetjie and Soetwater (Fig. 7). Hout Bay, Kommetjie and Soetwater show similarities in their stipes lengths to Buffels Bay, Batsata Rock and Betty’s Bay, and are significantly different to west of Cape Point counterpart sites such as Oudekraal and Scarborough. A difference for the majority of the morphologies are seen between Soetwater and Scarborough, with significant differences for epiphyte length, frond length, frond mass, stipe length, stipe mass and total length.  A separation along the same stretch of coast (west of Cape Point) for stipe mass shows that Hout Bay, Kommetjie and Soetwater are again significantly different to neighboring west side sites such as Oudekraal and Scarborough.

@@

When comparing morphologies bewteen deep and shallow water _Ecklonia maxima_ individuals, clear pattern is seen in certian morphological characteristics. These are frond length, frond mass, primary length, primary width and stipe circumference, and all show a similar pattern. Significant differences in these morphological characterstics at Kommetjie and Soetwater are evident, however there are no significant differences from Buffel's Bay to Kalk Bay.

@@

There are various strong correlations within the temperature parameters, and within the wave parameters on an annual time scale (Fig. 8). SD Significant wave height (Hs) and mean Hs showed a strong positive correlate(0.92), as well as  SD wave period (Tp) and mean Hs with a strong negative correlation(-0.9) For SD wind speed and mean wind speed a strong positive correlation existed (0.967). Minimum and mean temperatures correlate strongly (0.918), as well as SD wind speed and median wind direction (0.999). A strong correlation exists between SD temperature and temperature range (0.942). Median wind direction and mean wind speed are also correlated well (0.954). We however see no strong correlations between wave and temperature parameters.

These correlations vary through the three timescales (Annual, August and February). Although there are no strong correlations between temperature and wave parameters, we see interesting differences between the seasonal timescales The mean temperature and mean wind speed are weakly correlated in August (0.253)(Fig. 9), but are observed to be more strongly negatively correlated in February (-0.788)(Fig. 10).  The same is seen for minimum temperature and median wind direction in August (0.129) compared to February (-0.715).

@@

The morphometrics of _Ecklonia maxima_ were explained more by wave parameters (75%), than temperature parameters (66%), when separate RDA’s were constructed (Fig. 11).  The first two axes for wave parameters driving _Ecklonia maxima_ morphology explained 34% of the variation. Stipe circumference was positively influenced by both mean and SD annual significant wave height as well as annual mean wave direction. There was also a negative influence by annual SD of wave period and wave direction (which correlate with one another) on stipe circumference. The primary length of _Ecklonia maxima_ was influenced positively by annual mean wind direction, and negatively explained by annual SD wind direction.

@@

Although temperature parameters for _Ecklonia maxima_ showed some influence of the morphometric data, the explanation of morphology was not strong, with an R^2^ value of 0.662 and a negative adjusted R^2^ value (-0.138)(Fig. 12). Adjusted R^2^ values are modified R^2^ values to include the number predictors in a model. A negative adjusted R squared values equate to low explanatory power.  This is supported by the lack of relationships between environmental (temperature) and response (morphology) variables (Fig. 12).

@@

For _Laminaria pallida_, the percentage of explained data by wave parameters was 89%, compared to temperature parameters explaining 82% (Fig. 13). By using adjusted R^2^ values, the first two axes for wave parameters driving _Laminaria pallida_ morphology explained 57% of the variation of morphology in this species. An increase in annual SD of wave period and wave direction was a strong influence on lamina thickness of _Laminaria pallida_. Annual mean wave period (which negatively correlates with both annual SD of wave period and wave direction) saw a strong positive influence on lamina length, and a negative influence on lamina thickness. Stipe mass of _Laminaria pallida_ was influenced by annual mean wind direction, as well as annual mean significant wave height, wave direction and SD of significant wave height. The Total length of _Laminaria pallida_ specimens were similarly explained by annual mean significant wave height, wave direction and SD of significant wave height.

@@

Provided by adjusted R^2^ values, the first two axes of temperature effect on _Laminaria pallida_ morphologies amounted to only 11%, but exhibited a stronger influence than on _Ecklonia maxima_ (Fig. 14). The mean and minimum temperatures during February were observed to negatively influence _Laminaria pallida_ morphologies such as stipe mass, stipe length and stipe diameter, as well as thallus mass and total length. Maximum august temperatures however were observed to positively explain stipe diameter of _Laminaria pallida_.

@@

Seasonal variations in significant wave height (Hs), wave period and wave direction were observed from the data (Fig. 4, 5). The direction of swell swings to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006]. These swells were found to correlate with increased wave period (Tp), wave direction and Hs. False Bay is therefore shielded by these south westerly swells by the Cape Peninsula [@Shipley1964]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an explanation for increased variability of Hs and Tp. The wind speed and wind direction support the presence of upwelling in summer, where southerly winds blow parallel to the coast and trigger upwelling [@Field1980a]. This is supported by the decreased in temperature in summer (Fig. 2, 3), specifically along the west side of Cape Point. We however see a differentiation of median wind directions in summer, with the western coast of False Bay experiencing south easterly winds, compared to the west side of Cape Point experiencing south westerly winds. It is hypothesised that the topography and elevation along the Cape Peninsula channels, shields winds along the strip of land. This is however absent in winter, where strong northerly winds are experiences from St. Helena Bay to Betty’s Bay [@Field1980a]. St. Helena bay experiences decreased median wave directions, and is protected by a headland, similar to the sites found along the western side of False Bay, that are shielded by the Cape Peninsula. There is however increased variability in the wave direction from both False Bay sites and St. Helena Bay, which is encouraged by refraction of waves.

Morphological adaptation due to water motion may manifest itself in a number of ways to high wave energy environments. For instance, reduction of blade thickness, blade elongation, increase of stipe length, and stipe circumference increase and force of attachment. Although this study did not measure force of attachment, other morphological responses to wave parameters was evident. An increase in mean annual wave period saw an increase of lamina length of _Laminaria pallida_, but a decrease in lamina thickness (Fig. 13). By decreasing the thickness of the lamina, and increasing the length of the lamina (directly increasing surface area), a larger more flexible kelp can survive in environments with greater wave period. Increased thickness of cortical tissues within the lamina aid photosynthetic ability [@Aminina2012], however increased wave exposure deceases the boundary diffusion layer [@Raven1981; @Hurd2011]. Therefore _Laminaria pallida_ in increased wave period sites may be able to reduce the need for thick lamina for photosynthesis, as the increased wave period provide longer wave events, which decreases the diffusion boundary layer and allow easier nutrient uptake. Conversely we see that sites with reduced wave height and wave period, such as Baboon Rock, Miller’s Point, A-Frame and Roman Rock displayed the greatest lamina thickness, with better photosynthetic ability, for _Laminaria pallida_. Pace [@Pace1972] however studied the effect of wave exposure on _Macrocystis pyrifera_ morphologies, and found that lamina thickness had a strong positive relationship with wave exposure. This may be species specific, with _Macrocystis pyrifera_ attaining larger sizes, but with incomparable stipe and lamina morphologies to _Ecklonia maxima_ and _Laminaria pallida_.

Lamina thickness showed positive correlation with annual SD wave period (Fig. 13). Although this variable may just negatively correlate with annual mean wave period, another reason may exist. The greatest variation of wave period and wave direction occurs in August along the western side of False Bay. Kelps at these sites may have developed thicker lamina, with a possible strategy of maximising photosynthesis at the expense of maybe being dislodged through rare increased wave energy events. The ‘spreading out’ of the lamina may also suggest a ‘go with the flow’ tactic, to reduce breaking by being less stiff [@Friedland1995]. @Denny1997 explored the stress forces that wave period played on _Nereocystis luetkeana_ by exposing various kelp plant sizes to a range of wave period that would be found in coastal environments. Results showed that for shorter kelps (smaller stipe lengths), the maximum stress on a plant occurs at 10s wave periods, but only at 5.5s for larger kelps that are similar in length to the water depth.  Similar results were observed for kelps with larger blades, with maximal stress forces occurring at approximately 8s, compared to a stress ratio of only 0.4 at 16s wave periods.

Mean annual significant wave height (Hs) was observed to influence _Ecklonia maxima_ stipe circumference, with increased stipe circumference at larger Hs (Fig. 11). An increase in Hs saw greater wave actions as waves are larger, as well as greater wave periods that are positively correlated with Hs. An increase in stipe circumference provides _Ecklonia maxima_ with a more rigid structure to withstand the onslaught of waves.  These greater stipe circumferences of _Ecklonia maxima_ were observed at west side sites such as Soetwater and Kommetjie. While these sites experience increased Hs, they also are subject to seasonal upwelling events [@Andrews1980], that bring with it cold clean water. _Ecklonia maxima_ are thought to be influenced greatly by light attenuation [@Rothman2017], and thus have developed increased (stronger) stipe circumferences to be able to cope in large Hs environments, to access cleaner, nutrient-filled water. The same patterns are observed in _Laminaria pallida_ morphologies, that display increased stipe diameters at Soetwater and Kommetjie as well (Fig. 13).  This increase in stipe diameter correlates with stipe mass of _Laminaria pallida_, providing evidence that this species has developed thicker, strong stipes to persist in this region. Denny et al. (1997) found that the stress ratio decreased as stipe lengths increases, in a fixed depth. Therefore adult _Ecklonia maxima_ sporophytes would experience a lower stress ratio compared to adult _Laminaria pallida_ sporophytes in the same environment, where _Laminaria pallida_ is often a subsurface species to _Ecklonia maxima_.

Stipe circumference for _Ecklonia maxima_ were explained by annual wave direction (Fig. 11).  For _Ecklonia maxima_, the sporophytes may be producing stipes that have a greater circumference to cope with increased wave energy. The west side of the peninsula is more exposed to significant wave heights during winter due to the direction of swell, while other sites such as Betty's Bay and Basata rock are more protected due to their location along the coast and the presence of headlands which refract wave parameters.

Temperature parameters were not found to be a significant driver of kelp morphology in this study. Although temperature plays an important role in distribution of kelp and physiological functioning of adults and gametophytes, there is little evidence that temperature is a driver of morphological variation in kelps. Although turbidity is not a direct wave or temperature parameter, turbidity is influence by various abiotic variables. Increased swells around sandy beaches increase sedimentation and turbidity [@Yang2005], however wind direction and speed can encourage upwelling, with brings to the surface cold, clean, nutrient-rich water [@Boyd1998]. Increased wind speed at sites along the west side of the peninsula in a southerly direction trigger upwelling events [@Boyd1998]. We however see a lack of upwelling events for sites such as Betty’s Bay and Batsata Rock. These two sites display _Ecklonia maxima_ sporophytes with the largest total lengths (Fig. 7), as well as Betty’s Bay displaying the largest total length of _Laminaria pallida_ sporophytes(Fig. 6). The effect of depth on stipe length is regarded as insignificant in this study as all kelp were sampled at similar depths. Stipe elongation has been shown to be a morphological adaptation in wave exposed environment, as a longer kelp has time to extend back and forth in swell reducing the tension on the holdfast. Stipe lengths of _Ecklonia maxima_ from Kommetjie, Hout Bay and Soetwater are similar to Batsata rock and Betty’s Bay, but differs to west side neighbours such as Oudekraal and Scarborough (Fig. 7). Oudekraal and Scarborough experience more variability in wave direction, however they are observed to experience smaller annual mean significant wave heights, when compared to Kommetjie, Hout Bay, Soetwater, as well as Batsata Rock and Betty’s Bay (Fig. 4).@Friedland1995 found that longer kelps were able to reduce drag and tension as the sinusoidal displacement of most waves would not ‘stretch’ a kelp out long enough before falling. The ability to grow longer stipes would provide the kelp with a structure that could surpass most wave heights without the risk of dislodgement.

>>>>

### Background

@@

In Figure 3 the propagation of a typical offshore wave spectrum is given as produced from a single time-step in SWAN. Figure 3 is presented to clarify the averaged wave exposure maps presented in Figure 4. Tracing the wave height contours into False Bay its clear why this Bay’s western periphery is predominantly sheltered. It should be mentioned that some of the annual winter frontal depression systems pass the Cape Peninsula from the west to east, resulting in wave propagating towards the continent from much more southerly directions. This results in positive and negative wave exposure anomalies all around the peninsula.

@@

The temperature parameters around the Western Cape coast vary among sites and seasons. During February and August, the west side of Cape point experiences decreased temperatures (Fig. 5).  This region also experiences lower mean and maximum temperatures when compared to False Bay, a region known as east of Cape Point. The range of temperatures within False Bay are larger in winter (August) compared to summer (February) (Fig. 5). The sites north of Kommetjie on the west side display larger temperatures ranges than sites south of Kommetjie on the west side.

@@

Not only do the temperature parameters display seasonal fluctuations, but we see similar observations for wave parameters. The west side of Cape Point exhibit increased median direction, SD and mean significant wave height compared to within False Bay (Fig. 4). Sites within False Bay however display increased standard deviations of wave period. Median wind direction is northerly in winter (Fig. 6), and turns southerly in summer.

@@

#### _Laminaria pallida_
Lamina length for laminaria pallida showed no geographical pattern moving from west to east. Kommetjie, Olifantsbos and Batsata Rock showed great variability in lamina length, and both Buffels Bay and Betty’s Bay were visually different to Miller’s Point and Roman Rock, when comparing summary data (Fig.8). Lamina thickness showed great variation across sites, with Baboon Rock, Miller’s Point, A-Frame and Roman rock displaying large lamina thickness, significantly different to the rest of the sites. Lamina weight was observed to vary for Baboon Rock and Betty’s Bay. Neighbouring sites A-Frame and Roman Rock also showed visual difference when comparing boxplot summary statistics. An increase in the number of digits was observed as one moved from Cape Point north along the western side of False Bay. This ceased at Batstata Rock, which exhibited significantly less digits compared to the previous site, Bordjies reef North. Stipe diameter showed some geographical grouping, with west of Cape Point sites exhibiting larger stipe diameters compared to False Bay sites.

@@

For frond length, frond mass, stipe circumferences, stipe length and total length of _Ecklonia maxima_ morphometrics, we see a gradual increase in value as one moves south from St. Helena Bay to Kommetjie and Soetwater (Fig. 9). Hout Bay, Kommetjie and Soetwater show similarities in their stipes lengths to Buffels Bay, Batsata Rock and Betty’s Bay, and are significantly different to west of Cape Point counterpart sites such as Oudekraal and Scarborough. A difference for the majority of the morphologies are seen between Soetwater and Scarborough, with significant differences for epiphyte length, frond length, frond mass, stipe length, stipe mass and total length.  A separation along the same stretch of coast (west of Cape Point) for stipe mass shows that Hout Bay, Kommetjie and Soetwater are again significantly different to neighboring west side sites such as Oudekraal and Scarborough.

@@

When comparing morphologies bewteen deep and shallow water _Ecklonia maxima_ individuals, clear pattern is seen in certian morphological characteristics (Fig. 10). These are frond length, frond mass, primary length, primary width and stipe circumference, and all show a similar pattern. Significant differences in these morphological characterstics at Kommetjie and Soetwater are evident, however there are no significant differences from Buffel's Bay to Kalk Bay. <!-- I cannot get the black and white them to work on this plot...apologies I will try fix it as soon as possible. -->

@@

There are various strong correlations within the temperature parameters, and within the wave parameters on an annual time scale (Fig. 11). SD Significant wave height (Hs) and mean Hs showed a strong positive correlate(0.92), as well as  SD wave period (Tp) and mean Hs with a strong negative correlation(-0.9) For SD wind speed and mean wind speed a strong positive correlation existed (0.967). Minimum and mean temperatures correlate strongly (0.918), as well as SD wind speed and median wind direction (0.999). A strong correlation exists between SD temperature and temperature range (0.942). Median wind direction and mean wind speed are also correlated well (0.954). We however see no strong correlations between wave and temperature parameters (Fig 12).

These correlations vary through the three timescales (Annual, August and February). Although there are no strong correlations between temperature and wave parameters, we see interesting differences between the seasonal timescales. The mean temperature and mean wind speed are weakly correlated in August (0.253)(Fig. 12), but are observed to be more strongly negatively correlated in February (-0.788)(Fig. 13).  The same is seen for minimum temperature and median wind direction in August (0.129) compared to February (-0.715).

@@

The morphometrics of _Ecklonia maxima_ were explained more by wave parameters (75%), than temperature parameters (66%), when separate RDA’s were constructed (Fig. 14).  The first two axes for wave parameters driving _Ecklonia maxima_ morphology explained 34% of the variation. Stipe circumference was positively influenced by both mean and SD annual significant wave height as well as annual mean wave direction. There was also a negative influence by annual SD of wave period and wave direction (which correlate with one another) on stipe circumference. The primary length of _Ecklonia maxima_ was influenced positively by annual mean wind direction, and negatively explained by annual SD wind direction.

@@

Although temperature parameters for _Ecklonia maxima_ showed some influence of the morphometric data, the explanation of morphology was not strong, with an R^2^ value of 0.662 and a negative adjusted R^2^ value (-0.138)(Fig. 15). Adjusted R^2^ values are modified R^2^ values to include the number predictors in a model. A negative adjusted R squared values equate to low explanatory power.  This is supported by the lack of relationships between environmental (temperature) and response (morphology) variables (Fig. 15).

@@

For _Laminaria pallida_, the percentage of explained data by wave parameters was 89%, compared to temperature parameters explaining 82% (Fig. 16). By using adjusted R^2^ values, the first two axes for wave parameters driving _Laminaria pallida_ morphology explained 57% of the variation of morphology in this species. An increase in annual SD of wave period and wave direction was a strong influence on lamina thickness of _Laminaria pallida_. Annual mean wave period (which negatively correlates with both annual SD of wave period and wave direction) saw a strong positive influence on lamina length, and a negative influence on lamina thickness. Stipe mass of _Laminaria pallida_ was influenced by annual mean wind direction, as well as annual mean significant wave height, wave direction and SD of significant wave height. The Total length of _Laminaria pallida_ specimens were similarly explained by annual mean significant wave height, wave direction and SD of significant wave height.

@@

Provided by adjusted R^2^ values, the first two axes of temperature effect on _Laminaria pallida_ morphologies amounted to only 11%, but exhibited a stronger influence than on _Ecklonia maxima_ (Fig. 17). The mean and minimum temperatures during February were observed to negatively influence _Laminaria pallida_ morphologies such as stipe mass, stipe length and stipe diameter, as well as thallus mass and total length. Maximum august temperatures however were observed to positively explain stipe diameter of _Laminaria pallida_.

@@

Seasonal variations in significant wave height (Hs), wave period and wave direction were observed from the data. The direction of swell swings to the south west in winter, generated by strong low pressures that originate from the southern ocean [@Reason2006]. These swells were found to correlate with increased wave period (Tp), wave direction and Hs. False Bay is therefore shielded by these south westerly swells by the Cape Peninsula [@Shipley1964]. In summer, these swells rotate anticlockwise and are able to enter False Bay, providing an explanation for increased variability of Hs and Tp. The wind speed and wind direction support the presence of upwelling in summer, where southerly winds blow parallel to the coast and trigger upwelling [@Field1980a]. This is supported by the decreased in temperature in summer, specifically along the west side of Cape Point. We however see a differentiation of median wind directions in summer, with the western coast of False Bay experiencing south easterly winds, compared to the west side of Cape Point experiencing south westerly winds. It is hypothesised that the topography and elevation along the Cape Peninsula channels, shields winds along the strip of land. This is however absent in winter, where strong northerly winds are experiences from St. Helena Bay to Betty’s Bay [@Field1980a]. St. Helena bay experiences decreased median wave directions, and is protected by a headland, similar to the sites found along the western side of False Bay, that are shielded by the Cape Peninsula. There is however increased variability in the wave direction from both False Bay sites and St. Helena Bay, which is encouraged by refraction of waves.

Morphological adaptation due to water motion may manifest itself in a number of ways to high wave energy environments. For instance, reduction of blade thickness, blade elongation, increase of stipe length, and stipe circumference increase and force of attachment. Although this study did not measure force of attachment, other morphological responses to wave parameters was evident. An increase in mean annual wave period saw an increase of lamina length of _Laminaria pallida_, but a decrease in lamina thickness. By decreasing the thickness of the lamina, and increasing the length of the lamina (directly increasing surface area), a larger more flexible kelp can survive in environments with greater wave period. Increased thickness of cortical tissues within the lamina aid photosynthetic ability [@Aminina2012], however increased wave exposure deceases the boundary diffusion layer [@Raven1981; @Hurd2011]. Therefore _Laminaria pallida_ in increased wave period sites may be able to reduce the need for thick lamina for photosynthesis, as the increased wave period provide longer wave events, which decreases the diffusion boundary layer and allow easier nutrient uptake. Conversely we see that sites with reduced wave height and wave period, such as Baboon Rock, Miller’s Point, A-Frame and Roman Rock displayed the greatest lamina thickness, with better photosynthetic ability, for _Laminaria pallida_. Pace [@Pace1972] however studied the effect of wave exposure on _Macrocystis pyrifera_ morphologies, and found that lamina thickness had a strong positive relationship with wave exposure. This may be species specific, with _Macrocystis pyrifera_ attaining larger sizes, but with incomparable stipe and lamina morphologies to _Ecklonia maxima_ and _Laminaria pallida_.

Lamina thickness showed positive correlation with annual SD wave period. Although this variable may just negatively correlate with annual mean wave period, another reason may exist. The greatest variation of wave period and wave direction occurs in August along the western side of False Bay. Kelps at these sites may have developed thicker lamina, with a possible strategy of maximising photosynthesis at the expense of maybe being dislodged through rare increased wave energy events. The ‘spreading out’ of the lamina may also suggest a ‘go with the flow’ tactic, to reduce breaking by being less stiff [@Friedland1995]. @Denny1997 explored the stress forces that wave period played on _Nereocystis luetkeana_ by exposing various kelp plant sizes to a range of wave period that would be found in coastal environments. Results showed that for shorter kelps (smaller stipe lengths), the maximum stress on a plant occurs at 10s wave periods, but only at 5.5s for larger kelps that are similar in length to the water depth.  Similar results were observed for kelps with larger blades, with maximal stress forces occurring at approximately 8s, compared to a stress ratio of only 0.4 at 16s wave periods.

Mean annual significant wave height (Hs) was observed to influence _Ecklonia maxima_ stipe circumference, with increased stipe circumference at larger Hs. An increase in Hs saw greater wave actions as waves are larger, as well as greater wave periods that are positively correlated with Hs. An increase in stipe circumference provides _Ecklonia maxima_ with a more rigid structure to withstand the onslaught of waves.  These greater stipe circumferences of _Ecklonia maxima_ were observed at west side sites such as Soetwater and Kommetjie. While these sites experience increased Hs, they also are subject to seasonal upwelling events [@Andrews1980], that bring with it cold clean water. _Ecklonia maxima_ are thought to be influenced greatly by light attenuation [@Rothman2017], and thus have developed increased (stronger) stipe circumferences to be able to cope in large Hs environments, to access cleaner, nutrient-filled water. The same patterns are observed in _Laminaria pallida_ morphologies, that display increased stipe diameters at Soetwater and Kommetjie as well.  This increase in stipe diameter correlates with stipe mass of _Laminaria pallida_, providing evidence that this species has developed thicker, strong stipes to persist in this region. Denny et al. (1997) found that the stress ratio decreased as stipe lengths increases, in a fixed depth. Therefore adult _Ecklonia maxima_ sporophytes would experience a lower stress ratio compared to adult _Laminaria pallida_ sporophytes in the same environment, where _Laminaria pallida_ is often a subsurface species to _Ecklonia maxima_. Stipe circumference for _Ecklonia maxima_ were explained by annual wave direction.  For _Ecklonia maxima_, the sporophytes may be producing stipes that have a greater circumference to cope with increased wave energy. The west side of the peninsula is more exposed to significant wave heights during winter due to the direction of swell, while other sites such as Betty's Bay and Basata rock are more protected due to their location along the coast and the presence of headlands which refract wave parameters.

Temperature parameters were not found to be a significant driver of kelp morphology in this study. Although temperature plays an important role in distribution of kelp and physiological functioning of adults and gametophytes, there is little evidence that temperature is a driver of morphological variation in kelps. Although turbidity is not a direct wave or temperature parameter, turbidity is influence by various abiotic variables. Increased swells around sandy beaches increase sedimentation and turbidity [@Yang2005], however wind direction and speed can encourage upwelling, with brings to the surface cold, clean, nutrient-rich water [@Boyd1998]. Increased wind speed at sites along the west side of the peninsula in a southerly direction trigger upwelling events [@Boyd1998]. We however see a lack of upwelling events for sites such as Betty’s Bay and Batsata Rock. These two sites display _Ecklonia maxima_ sporophytes with the largest total lengths, as well as Betty’s Bay displaying the largest total length of _Laminaria pallida_ sporophytes. The effect of depth on stipe length is regarded as insignificant in this study as all kelp were sampled at similar depths. Stipe elongation has been shown to be a morphological adaptation in wave exposed environment, as a longer kelp has time to extend back and forth in swell reducing the tension on the holdfast. Stipe lengths of _Ecklonia maxima_ from Kommetjie, Hout Bay and Soetwater are similar to Batsata rock and Betty’s Bay, but differs to west side neighbours such as Oudekraal and Scarborough. Oudekraal and Scarborough experience more variability in wave direction, however they are observed to experience smaller annual mean significant wave heights, when compared to Kommetjie, Hout Bay, Soetwater, as well as Batsata Rock and Betty’s Bay. @Friedland1995 found that longer kelps were able to reduce drag and tension as the sinusoidal displacement of most waves would not ‘stretch’ a kelp out long enough before falling. The ability to grow longer stipes would provide the kelp with a structure that could surpass most wave heights without the risk of dislodgement.
